---
title: "Additional Exercises"
author: by Jeff Chen & Dan Hammer, Georgetown University McCourt School of Public
  Policy
subtitle: Intro to Data Science for Public Policy, Spring 2016
output:
  pdf_document:
    toc: yes
  html_document:
    theme: journal
    toc: yes
---

##Additional Exercises! + Answers below

1. Write a function "unique2" to calculate the number of unique values in the following series. The series below is normally distributed ( $n = 100$ with $\mu = 50$ and $\sigma = 10$), rounded to the nearest integer.
```{r}
  set.seed(123)
  x <- round(rnorm(100,50,10))
```


2. Imputation of missing values is often times involves replacing unknown values with the mean, median, or mode.  Mean can be calculated using `mean()`, median using `percentile()`, and mode using a combination of `table()`, `sort()`, `names()`. For categorical data, the mode is most appropriate. In the vectors below, replace all `NA` values with the most common string value in `y1` and the mean for the `y2`.
```{r}
  y1 <- c("a","a",NA,"c","d","e", NA,"f", NA,"g","a", NA, "a","c","z","g")
  y2 <- c(1, 10, NA, 5, 6, 1, NA, NA, 9, NA, 15, 3, NA, NA, 3, 9, 2)
```


3. Convert your imputation code into a function named `plug.it()` that accepts a vector of any kind. The logic of the function should be as follows:

- function(vec)
  - if class of vec is numeric, then impute with mean
  - if class of vec is character, factor, or logical, then impute with mode
  - return imputed set
  
Try out your function below.

```{r}
  y3 <- c("z","z",NA,"c","d","e", NA,"f", NA,"s","z", NA, "a","c","z","g")
  y4 <- c(2, 11, NA, 3, NA, 5, 6, 1, NA, NA, 9, NA, 16, 3, NA, NA, 2, 4, 3)
```

4. Write a loop to impute each column of the following data frame using `plug.it()`. Since your function is flexible, all you'd need to do is loop column 1 through 3.
```{r}
  df <- data.frame( x1 = c(2, 11, NA, 3, NA, 5, 6, 1, NA, NA, 9, NA, 16, 3, NA, NA, 2, 4, 3),
                    x2 = c(20, NA, NA, 30, NA, 15, 6, 1, 10, 11, 9, 2, 16, 3, 400, 500, 2, 4, 3),
                    x3 = c(228, NA, NA, 39, NA, 2, 6, 1, 2, 5, 3, 2, NA, 3, NA, NA, 5, 2, 34))
```


##Answers!
1. Answer. Steps:
- create a placeholder `p`
- loop through all values of vector `vec`
  - if index `i` is not in placeholder `p`, then append index `i` to end of `p`
- when done, return `p`
```{r}
  unique2 <- function(vec){
    #create placeholder
      p <- c()
      
    #loop
      for(i in vec){
        if(!(i %in% p)){
          p <- c(p, i)
        }
      }
    
    #return
      return(p)
  }

#test it
  set.seed(123)
  x <- round(rnorm(100,50,10))
  a <- unique2(x)
```

2. Answer! 
- Use `table()` to return a list of unique values with their frequencies
- Sort the table descending, extract the first value (letter)
- Replace all blanks with the letter
```{r}
#String 
  y1 <- c("a","a",NA,"c","d","e", NA,"f", NA,"g","a", NA, "a","c","z","g")
  tab <- table(y1)
  letter <- names(sort(tab,decreasing=TRUE))[1]
  y1[is.na(y1)] <- letter
  print(y1)

#Numeric 
  y2 <- c(1, 10, NA, 5, 6, 1, NA, NA, 9, NA, 15, 3, NA, NA, 3, 9, 2)
  mu <- mean(y2, na.rm = T)
  y2[is.na(y2)] <- mu
  print(y2)

```


3. Answer! 

```{r}
#Function
  plug.it <- function(vec){
    
    if(class(vec)=="numeric"){
      #numeric!
        mu <- mean(vec, na.rm = T)
        vec[is.na(vec)] <- mu
        
    } else if(class(vec) == "character" | class(vec) == "factor"){
      
      #string!
        tab <- table(vec)
        letter <- names(sort(tab,decreasing=TRUE))[1]
        vec[is.na(vec)] <- letter
    }
    return(vec)
  }

#Test
  y3 <- c("z","z",NA,"c","d","e", NA,"f", NA,"s","z", NA, "a","c","z","g")
  y4 <- c(2, 11, NA, 3, NA, 5, 6, 1, NA, NA, 9, NA, 16, 3, NA, NA, 2, 4, 3)
  
  plug.it(y3)
  plug.it(y4)

```


4. Answer! 

```{r}
  df <- data.frame( x1 = c(2, 11, NA, 3, NA, 5, 6, 1, NA, NA, 9, NA, 16, 3, NA, NA, 2, 4, 3),
                    x2 = c("a","a", NA,"c","d","e", NA,"f", NA,"g","a", NA, "a","c","z","g", "x", "g", "z"),
                    x3 = c(20, NA, NA, 30, NA, 15, 6, 1, 10, 11, 9, 2, 16, 3, 400, 500, 2, 4, 3),
                    x4 = c(228, NA, NA, 39, NA, 2, 6, 1, 2, 5, 3, 2, NA, 3, NA, NA, 5, 2, 34))
  #Loop it
  for(i in 1:ncol(df)){
    df[,i] <- plug.it(df[,i])
  }
  
  #view
  print(df)

```