---
title: "Lecture 2: Data Manipulation"
subtitle: "Intro to Data Science for Public Policy, Spring 2016"
author: "by Jeff Chen & Dan Hammer, Georgetown University McCourt School of Public Policy"
output: 
  html_document: 
    theme: journal
    toc: yes
---


In this course, we will maingly work with data sets that are rectangular.  Rows of the data table are analytical units, like countries or voters.  Columns are attributes of the units, like GDP or gender.  An example of a data table is voter composition by county, with information on population, ethnicity, and economic characteristics.  The county is the analytical unit and the metric of population is an attribute.  This example data represents a basis data set.  Analysis of the data set requires *views* onto the table -- pulling out subsets for analysis.

Suppose we wanted to rank all Michigan counties by share of minority voters.  Or suppose we wanted to aggregate the counties to find the U.S. state with the highest per capita income.  These *views* on the data table require operations to pivot, merge, aggregate, and sort the raw data.  

We will review the most powerful functions for data manipulation in the lecture: `sort`, `reshape`, `collapse`, and `merge`.  A mastery of these basic operations can yield just about any derived data set from a structured table in `R`.

### Indexing and Sorting

Consider a data table with two individuals, indexed by the variable `id`, over two time periods, indexed by the variable `t`.  Each individual has a different record for each time period, with observations on income (`income`) and voter preference (`vote`) on a 1-10 scale where higher numbers indicate more progressive voting.  Consider the data table stored in variable `X`:

```{r comment=NA, cache=TRUE}
(X <- data.frame(id=c(1,1,2,2), t=c(1,2,1,2), income=c(50,55,101,123), vote=c(8,7,4,3)))
```

##### Extract second record
```{r comment=NA, cache=TRUE}
X[2, ] # extract second row
```

##### Extract income for all records
```{r comment=NA, cache=TRUE}
X[, 3] # extract third column
X[, "income"] # extract column with "income" label
X[["income"]] # list syntax to extract column from data frame
X$income # compact version of data column manipulation
```

##### Extract multiple records
```{r comment=NA, cache=TRUE}
print(1:3) # index range
X[1:3, ] # apply the index range to extract rows
X[c(1, 3), ] # specific indices
```

##### Extract multiple columns
```{r comment=NA, cache=TRUE}
X[ , 3:4] # extract multiple columns
X[ , c("income", "vote")] # multiple column labels
```

##### Extract records by attribute value
```{r comment=NA, cache=TRUE}
(idx <- X[["income"]] > 50) # indices of records with income greater than 50
X[idx, ] # select the indices
```

##### Reorder records
```{r comment=NA, cache=TRUE}
X[c(2, 4, 3, 1), ] # random ordering, note row labels
X[c(4, 3, 2, 1), ] # backwards ordering (by force)
```

##### Ordering and sorting
```{r comment=NA, cache=TRUE}
order(X$vote)
X[order(X$vote), ] # order records by vote
X[order(-X$vote), ] # order records by vote, decreasing
X[order(X$vote, decreasing=TRUE), ] # order records by vote, decreasing
```

##### Order by multiple columns
```{r comment=NA, cache=TRUE}
X[order(X$t, X$vote), ] # order records, first by time period and then by vote
X[order(X$t, -X$vote), ] # order records, first by time period and then dec. by vote
```

#### Exercises 2.1

1. Extract records with income greater than 120 in the second time period
2. Extract records with vote greater than 5 and income less than 51
3. Briefly explain the difference between assigning a variable with surrounding parentheses versus no parentheses.  How does this behavior relate to the concept of a [side effect](https://en.wikipedia.org/wiki/Side_effect_(computer_science))?


### Reshape

There are two basic *shapes* of data, **wide** and **long**.  The data table in the previous section was **long**, where each row represents an individual in a separate time period.  Row 2 in the original table `X`, for example, was the first individual in the second time period:

```{r comment=NA, cache=TRUE}
X[2, c("id", "t")]
```

We will find that the **wide** format is also useful.  Here, the wide format is every attribute for each individual over time.  The analytical unit is the individual, not the individual in each, separate time period.  An attribute is no longer just *income* but rather *income in time period*.  

```{r comment=NA, cache=TRUE}
(wide <- reshape(X, idvar="id", timevar="t", direction="wide"))
```

There are now just two rows, where there were previously four.  Returning to the original **long** format is straightforward; but we aren't left with the exact same data table.  There are artifacts of the change-in-shape for both the column and row names.  Sort of like data manipulation breadcrumbs.  This will be cleaned up in the next subsection.

```{r comment=NA, cache=TRUE}
(long <- reshape(wide, idvar="id", timevar="t", direction="long"))
```

Note that there are other frameworks that may be more useful, depending on your application.  An especially useful framework to "melt" and "cast" data is the [`reshape`](http://w.statpower.net/Content/MLRM/Lab/Reshape%20Paper.pdf) package.

#### Rename row and column headers

The newly assigned column or row names may not match the meaning in the data.  It is good practice to maintain the column headers at each stage of analysis, even if you don't immediately use the intermediate data table.  Otherwise editing code gets confusing, quickly.  The column names are stored in an attribute of the data frame:

```{r comment=NA, cache=TRUE}
names(long)
```

Renaming column headers using the built-in, base functions in `R` looks confusing.  In words, the following code identifies the positions in `names(long)` where the values are `income.1` and `vote.1`.  At the specified positions, the values is reassigned with new values `income` and `vote`, respectively.

```{r comment=NA, cache=TRUE}
names(long)[names(long) == "income.1"] <- "income"
names(long)[names(long) == "vote.1"] <- "vote"
print(long)
```

This relatively simple example offers an overture to non-core `R` functions.  The standard `R` functions are extended through external packages.  For example, the `rename()` function is provided through the `plyr` package and abstracts away the list indexing idiosyncracies of renaming columns.  It packages the previous reassignments into more readable code:

```{r comment=NA, cache=TRUE, message=FALSE}
library(plyr)
long <- reshape(wide, idvar="id", timevar="t", direction="long")
rename(long, c("income.1"="income", "vote.1"="vote"))
```

### Collapse

Reshape and collapse are often used in conjunction in order to calculate summary statistics by group.  The built-in `R` functions are reasonably effective; but there are frameworks that are much more powerful.  Consider the [`reshape`](http://w.statpower.net/Content/MLRM/Lab/Reshape%20Paper.pdf) package for more complex data manipulation.  To round out the basic concept, consider the aggregate function to collapse functions by group.

Consider this function to collapse the data frame by `id` to create a new data frame with the average vote and income over time.

```{r comment=NA, cache=TRUE}
(aggdata <- aggregate(X, by=list(X$id), FUN=mean))
aggdata[c("id", "income", "vote")]
```

#### Exercises 2.2

1. Load the [`iris`](http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/iris.html) dataset using the `datasets` package.
2. Calculate the average and maximum sepal length for each Iris species. (**Bonus**: Write the commands without error messages.) 
3. Use the `aggregate()` function to count the number of observations of sepal width for each species.
4. (*Difficult*) Create a data frame with the 35th observation of sepal width for each species.

### Merge

Create two data frames for illustrative purposes, based on the [`state`](http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/state.html) datasets.  

```{r comment=NA, cache=TRUE}
library(datasets)
region <- data.frame(name=state.name, region=state.region)
area <- data.frame(name=state.name, area=state.area)
```

Suppose we want to merge the two datasets, so that we have a single, master dataset with the region and area of the states -- a fifty-by-three

```{r comment=NA, cache=TRUE}
full.data <- merge(region, area, by="name")
head(full.data)
```

Now we can see the value of the composition of these operations.  The following sequence identifies the region of the United States with the greatest land area:

```{r comment=NA, cache=TRUE}
df <- aggregate(full.data$area, by=list(full.data$region), FUN=sum)
df <- rename(df, c("Group.1"="region", "x"="total.area"))
df[order(df$total.area, decreasing=TRUE), ]
```

#### Exercises 2.3

1. Use the [`state`](http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/state.html) datasets to identify the state divisions (e.g., New England) with the highest murder rates in 1976.
2. What was the recorded population in 1977?



