---
title: "Lectures 3 + 4: Exploratory Data Analysis"
output: html_document
---

```{r, echo=FALSE, warnings = FALSE, messages=FALSE, id="asd"}
# Load package
library(networkD3)

# Create fake data
src <-    c("1 : investigate", "2 : ask","3 : design", "4 : collect", "5 : explore", "6 : reshape", "7 : model", "8 - communicate","9 - deploy", "5 : explore","5 : explore","5 : explore","5 : explore","5 : explore")
target <- c("2 : ask","3 : design", "4 : collect",  "5 : explore", "6 : reshape", "7 : model", "8 - communicate", "9 - deploy",   "10 - refine","central tendency","data distribution","outliers","patterns","data formats")
networkData <- data.frame(src, target)

# Plot
simpleNetwork(networkData,  charge = -500, opacity=1,linkDistance = 30, fontSize = 14,width='100%', height='300px', zoom=T)
```
 

Lecture 3/4: Exploratory Data Analysis
detection of mistakes • checking of assumptions • preliminary selection of appropriate models • determining relationships among the explanatory variables, and • assessing the direction and rough size of relationships between explanatory and outcome variables.

Univariate EDA
Categorical or Discrete Data
- Tab
- Prop.table

Numerical Data
- summary()
- mean()
- median()
- sd()
- var()
- range()
- IQR()
- quantile()

Graphical
- ggplot examples: http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/
- par(mfrow = c(2,1)) 
    - hist()
- boxplot
    - boxplot(numeric~group, data=df, main=“Title”, 
    -    xlab="Number of Cylinders", ylab="Miles Per Gallon")
    - boxplot(len~supp*dose, data=ToothGrowth, notch=TRUE, 
    -   col=(c("gold","darkgreen")),
    -   main="Tooth Growth", xlab="Suppliment and Dose")
- 2-way relationships
    - scatter
    - scatterplot matrix 
- High density
    - scatter with density
- Correlation matrix
- 

Bivariate
- Correlation 
    - corrplot
- Covariance



```{r, warning=FALSE, message=FALSE}
##QQPLOT
par(mfrow=c(2,2))

for(k in c(5,20,100,100)){
  pois = rpois(k,lambda=21)
  norm = rnorm(length(pois), mean(pois), sd(pois))
  qqplot(norm,pois, main=paste0("n=",k), aspect=1)
  lines(norm,norm)
}



```
