 ---
title: "Lecture 3: Data Quality "
output: html_document
---

## Lecture 3: Data Quality

The quality of data can be a selling point or achilles heel of a data science project.

Checking data quality
- Missingness
- Standardized values
- Visual checks

The objective of this lecture is to handle missing values appropriately and script visual checks to find errors introduced in data input/output. We will also start to view computational optimization techniques, like taking advantage of multiple cores for heavy duty operations (parallel processing).

##### Create a random dataset
Consider a dataset of *n* individuals from a  community. The data follows individuals over a 100-month period. 
```{r}

n = 10000

set.seed(123)

##Create main dataset with age, education and gender
df <- data.frame(id=seq(1,n,1), 
                age = sample(25:85,n, replace=T), 
                education = sample(c("High School","H.S.","High School","High School","HS","College","Bachelor","Bachelor's","Master's",NA),n,replace=T),
                gender = sample(c("M","F","-999"),n,replace=T)
                )

##Simulate missing data. Add 100 columns (1 to 100) of 
##randomly generated numbers using runif() that represent follow up data collection
##Non-response (-999) grows in a sigmoid fashion over time.
for(k in 1:100){
  
  #Each month is returned as a vector, drawn from 
  # a distribution of randomly generated numbers
  # where the occurence of non-response (-999) grows over time
  temp = sample(c(rep(-999,k^2),
           runif(round(10000/k),400,10000)),10000, replace=TRUE)
  
  #Vector is assigned a name in sequence ("inc1", "inc2",...)
  assign(paste0("inc",k),temp)
  
  #Result is column binded to the main dataset
  df <- cbind(df,mget(paste0("inc",k)))
  
}


```

##### Examine dataset
```{r}
#Retrieve first few rows
  head(df,1)

#Obtain structure of dataset
  str(df)

#obtain summary for first 10 variables
  summary(df[,1:10]) 

```


##### Check for steps in dataset
```{r}
#Check education
  table(df$education)
  table(df$education, useNA = "always")
  unique(df$education)

#Check gender
  table(df$gender)
  table(df$gender, useNA = "always")
```


##### Standardize levels in a dataset
```{r}

str(df$education)

hs <- c("H.S.","High School","HS")
college <- c("Bachelor","Bachelor's","College")

levels(df$education)[levels(df$education) %in% hs ] <- "High School"
levels(df$education)[levels(df$education) %in% college ] <- "College"
levels(df$education)[levels(df$education)=="Master's"] <- "Graduate"
 
```


##### Calculate missingness over time

We'll now write a simple function to return the proportion of records that are missing responses (-999) for a given month in the dataset. This new function 'missing' takes the following arguments:

- k: index number of a column
- dataframe: name of dataframe

```{r}
  missing <- function(k){
    mean(df[,k]==-999)
  }
```

To test this, let's look at column 95:

```{r}
  missing(95)
```

To put this to the test, we'll now put this to the test. Using lapply(), we can loop through all months of data (columns 5 to 104) and return a list containing the percent missing per month. 

```{r}
count <- lapply(5:ncol(df), FUN = missing)

```

Since lapply returns a list object, we'll need to apply unlist() to convert the object into a vector, then add it into a data frame with row labels for each month.

```{r}
count <- data.frame(month = 1:100, pct_missing = unlist(count))
plot(count)
```


###Exercise 3.1: 
GHCN-M: Missing values analysis in matrix of weather anomalies from 1880 to Present
Get data: [NOAA GHCN-M](https://www.kaggle.com/noaa/global-historical-climatology-network)

```{r}

```

Weather: Matching human-reported weather events to radar events over time and space

