---
title: "Lecture 1: Preliminaries"
subtitle: "Intro to Data Science for Public Policy, Spring 2016"
author: "by Jeff Chen & Dan Hammer, Georgetown University McCourt School of Public Policy"
output: 
  html_document: 
    theme: journal
    toc: yes
---


The objective of the first lecture is to overcome the coefficient of static friction in using R for data science. Students will learn to execute simple R scripts to read, write, and extract data elements.

##Overview
###Data Science and its many possibilities
- Computer vision 
- Language
- Forecasting
- Scoring



###What is data science?
[definition goes here]
- Comms



##Software Tools and Code Languages

Common languages 
- R, Python, Julia, Jav

###Installing RStudio
[screen shots]
- IDE

```{r,  message=FALSE, warning=FALSE}
#
 
```

###Getting started with Github
Getting started with Github

[text goes here]
```{r,  message=FALSE, warning=FALSE}
#
 
```

##Get started with R
###Read in data
[Basic operations]
- Example of solar energy production

The amount of sunlight impacting the array during the past minute.  

```{r,  message=FALSE, warning=FALSE}
df <- read.csv("https://s3.amazonaws.com/net-zero/2015-data-files/PV-hour.csv")
df$Timestamp <- strptime(df$Timestamp, "%Y-%m-%d %H:%M:%S")
df$month <- format(df$Timestamp, "%m")

library(ggplot2)
ggplot(df, aes(factor(month),PV_PVInsolationHArray)) + geom_boxplot(colour = "grey", fill = "navy")  + ggtitle("Boxplot") + labs(x = "Period")

```


###Data Formats
- Data formats map to different fields

| **Abbreviation**  | **Name + Description** | **Type** |**Example** | **Fields of Use ** | 
|-------+---------------------+---------+--------------------------------------+--------------------|-------|
| CSV | Comma Separated Values | Tabular | state, pop_mil <br> California, 38.8 <br> Delaware, 0.936 | Widely used format in most fields|
| TSV | Tab Separated Values | Tabular | state    pop_mil <br> California    38.8 <br> Delaware    0.936 | Widely used format in most fields|
| JSON  | JavaScript Object Notation | Tabular | [{"state": "California", "pop_mil": 38.8},<br>{"state": "Delaware", "pop_mil": 0.936}] | A common format provided from APIs |
| XML | Extensible Markup Language | Tabular | \<data><br>&nbsp;&nbsp; \<rec> <br> &nbsp; &nbsp;&nbsp;   \<state>California \</state><br> &nbsp; &nbsp;&nbsp;   \<pop_mil>38.8 \<pop_mil><br> &nbsp;&nbsp;\</rec> <br> &nbsp;&nbsp; \<rec> <br> &nbsp; &nbsp;&nbsp;   \<state>Delaware \</state><br> &nbsp; &nbsp;&nbsp;   \<pop_mil>0.936 \<pop_mil><br> &nbsp;&nbsp;\</rec><br> \<\data>| Widely used format in most fields|
| SHP | Shapefile | Spatial | Contains four files: .shp, .shx, .sbn, .dbf   | Common format for GIS |
| GeoJSON | GeoJSON | Spatial  |   | Common format for web-based GIS |
| TIFF | Tagged Image File Format | Spatial + Imagery  |   | Common format for representing grid-based|
| HDF | Hierarchical Data Format | Spatial + Imagery  |   | Common format for representing grid-based|
| NetCDF | Network Common Data Form (NetCDF) for sharing array-oriented scientific data | Spatial + Imagery  |   | Common format for representing grid-based|

##Example
###Example application
```{r}

```
###Exercise Data
-Health from the Demographic and Health Surveys or Census data.
